---
title: "Assignment 3"
author: "Arturo Arranz Mateo"
date: "March 30, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question1

You can also embed plots, for example:

```{r plotting, echo=FALSE}
#Data 
rm(list=ls())
mydata = read.csv("/home/arturo/Documents/Time Series Analysis/Assignment3/A3_jagt_NOx.csv")

#plot
str(mydata)
mydata$hour <-substr(as.character(mydata$Time),2,3)
mydata$ctime <- as.POSIXlt(paste(as.character(mydata$Date),mydata$hour),format="%d-%m %H")
trainData = mydata[1:696,]
testData = mydata[697:744,]

plot(trainData$ctime, trainData$NOx, type="l", col = "blue")


(arima000 <- arima(trainData$NOx, c(0, 1, 0),seasonal = list(order = c(0, 1, 1), period = 24)))
plot(trainData$ctime ,arima000$residuals, type="l", col = "green")
acf(arima000$residuals)
pacf(arima000$residuals)

prediction = predict(arima000,48)

plot(testData$ctime, testData$NOx, ylim = c(-100,200))


lines(testData$ctime ,prediction$pred, col="green")
lines(testData$ctime ,prediction$pred-prediction$se, col="red")
lines(testData$ctime ,prediction$pred+prediction$se, col="red")

```
From the raw data plotting we can infer the following observations:
- There is not a general trend
- Looks like there is a 24 hours seasonality

If we want to look closer, we can subplot each week
```{r plotting1, echo=FALSE}
plot(mydata$ctime[1:168], mydata$NOx[1:168], type="l", col = "blue")
plot(mydata$ctime[168:336], mydata$NOx[168:336], type="l", col = "blue")
plot(mydata$ctime[336:504], mydata$NOx[336:504], type="l", col = "blue")
plot(mydata$ctime[504:744], mydata$NOx[504:744], type="l", col = "blue")
```
Here the seasonality and day-trend is more obvious. However, we can see that the 3th week data is far from stationary.

We can applay some transformations, like one time differentation:
```{r plotting2, echo=FALSE}
plot(mydata$ctime[337:504], diff(mydata$NOx[336:504]), type="l", col = "blue")
```
Now we can see that the mean is constant, and we have remove the trend.
However, the variance do not seem constant throught all the data.

We can applay a logarithmic transformation before differencing.
```{r plotting6, echo=FALSE}
plot(mydata$ctime[337:504], diff(log(mydata$NOx[336:504])), type="l", col = "blue")
#qqnorm(diff(log(mydata$NOx)))
#qqline(diff(log(mydata$NOx)))

#qqnorm(diff(log(mydata$NOx),6));
#qqline(diff(log(mydata$NOx),6));
```
Now the data seems much more stationary. However we still can clearly see a strong correlation between consecutive samples. For filtering that correaltion we will use the ARMA model.

Here is how the whole data looks after both transformations:
```{r plotting5, echo=FALSE}
plot(mydata$ctime[-1], diff(log(mydata$NOx)), type="l", col = "blue")
```


## Question2

ACF and PACF
```{r pacf1, echo=FALSE}
acf(trainData$NOx)
pacf(trainData$NOx)
```

ACF and PACF with log and differentation transformations
```{r pacf2, echo=FALSE}
acf(diff(log(trainData$NOx)))
pacf(diff(log(trainData$NOx)))
```


## Question3
```{r pressure, echo=FALSE}
(arima111 <- arima(log(trainData$NOx), c(1, 1, 2),seasonal = list(order = c(0, 1, 0), period = 24)))
(arima211 <- arima(log(trainData$NOx), c(1, 1, 1),seasonal = list(order = c(1, 1, 1), period = 24)))
(arima110 <- arima(log(trainData$NOx), c(1, 1, 0),seasonal = list(order = c(1, 1, 0), period = 24)))
(arima210 <- arima(log(trainData$NOx), c(2, 1, 0),seasonal = list(order = c(2, 1, 0), period = 24)))
predictions = predict(arima111,48)
plot(trainData$ctime ,arima111$residuals, type="l", col = "green")


qqnorm(arima112$residuals)
qqline(arima112$residuals)
acf(arima112$residuals)
pacf(arima112$residuals)

qqnorm(arima211$residuals)
qqline(arima211$residuals)
qqnorm(arima110$residuals)
qqline(arima110$residuals)
qqnorm(arima210$residuals)
qqline(arima210$residuals)

plot(testData$ctime, testData$NOx)
lines(testData$ctime ,2.718^predict(arima000,48)$pred, col="green")
lines(testData$ctime ,2.718^predict(arima211,48)$pred, col="red")
lines(testData$ctime ,2.718^predict(arima110,48)$pred, col="blue")
lines(testData$ctime ,2.718^predict(arima210,48)$pred, col="pink")



NOx <- as.matrix(mydata[3])
tday <- seq(from=24, to=1)
NOxday <- matrix(NOx, ncol = 24, byrow = TRUE) #reshape into day rows



(fit <- arima(log(NOx), c(0, 1, 1),seasonal = list(order = c(0, 1, 1), period = 24)))
predict(fit,10)
plot(mydata$NOx, type="l")
#First day plot
plot(tday, NOxday[1,])
plot(tday, NOxday[31,])
plot(tday, NOxday[2,])
plot(tday, NOxday[,])
plot(tday, NOxday[5,])
plot(tday, NOxday[6,])

acf(log(NOx))
pacf(log(NOx))

acf(NOx)
pacf(NOx)

acf(diff(NOx,lag=1))
pacf(diff(NOx,lag=1))



pred <- predict(fit, n.ahead = 2*24)
ts.plot(NOx,2.718^pred$pred, log = "y", lty = c(1,3))

time <- ts(1:length(mydata$NOx), frequency = 24)
#The complete serie plotting
plot(time, mydata$NOx,col="green",type="l", lty=2, lwd=1)
plot(time, log(NOx),col="green",type="l", lty=2, lwd=1)
plot(diff(NOx,lag=1),col="green",type="l", lty=2, lwd=1)

dates <- seq(as.Date("14/03", format = "%d/%m"),by = "days", length = length(mydata$NOx)/24)
plot(time,NOx,col="green",type="l", lty=2, lwd=1)
axis(1, at=seq(1, length(mydata$NOx), by = 24), labels=dates)
require(ggplot2)
theme_set(theme_bw()) # Change the theme to my preference
ggplot(aes(x = time, y = variable), data = mydata) + geom_point()


t = seq(length(NOx))
plot(timets,NOx)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
